---
title: "Single Variable Models"
output: github_document
---

The purpose of this notebook is to explore single variable models and determine which features correlate best with the likelihood of churning. The first thing we'll need to do is gather the data from the `data` directory.

```{r include = FALSE, warning = FALSE, message = FALSE}
# load libraries
library(dplyr)
library(ggplot2)
library(ROCR)
```

```{r}
# read data from csv
subs <- read.csv('../data/features.csv', header = T)
```

Great, we have 74 thousand subscriptions to work with. To evaluate our models, we'll need to split our data into a training and testing set.

```{r}
# set seed for reproducible results
set.seed(2356)

# set random groups
subs$rgroup <- runif(dim(subs)[[1]])

# split out training and testing sets
training <- subset(subs, rgroup <= 0.8)
testing <- subset(subs, rgroup > 0.8)
```

Great, 80% of our data was put into a training set and the remaining 20% in a testing set. 

### Single Variable Models
Next we'll need to define some functions that will help us make predictions and calculate area under the curve (AUC). The code has been hidded but is available in the .Rmd file.

```{r warning = F, message = F, include = F}
# Given a vector of outcomes (outCol), a categorical training variable (varCol), 
# and a prediction variable (appCol), use outCol and varCol to build a single-variable model 
# and then apply the model to appCol to get new predictions.

pos <- 1

make_prediction_categorical <- function(outCol, varCol, appCol) {
  
  # Find how often the outcome is positive during training
  pPos <- sum(outCol == pos) / length(outCol)
  
  # We need this to handle NA values
  naTab <- table(as.factor(outCol[is.na(varCol)]))
  
  # Get stats on how often outcome is positive for NA values in training
  pPosWna <- (naTab/sum(naTab))[pos]
  
  vTab <- table(as.factor(outCol),varCol)
  
  # Get stats on how often outcome is positive, conditioned on levels of the variable
  pPosWv <- (vTab[pos, ] + 1.0e-3 * pPos) / (colSums(vTab) + 1.0e-3)
  
  # Make predictions by looking up levels of appCol
  pred <- pPosWv[appCol]
  
  # Add predictions for NA values
  pred[is.na(appCol)] <- pPosWna
  
  # Add in predictions for levels of appCol that weren’t known during training
  pred[is.na(pred)] <- pPos
  
  pred

} 


# function to calculate AUC
calcAUC <- function(predcol, outcol) {
  
  perf <- performance(prediction(predcol, outcol == pos), 'auc') 
  as.numeric(perf@y.values)
  
}

# function that makes predictions for numeric variables
make_prediction_numeric <- function(outCol, varCol, appCol) {
  
  # Make the cuts to bin the data
  cuts <- unique(as.numeric(quantile(varCol, probs = seq(0, 1, 0.1), na.rm = T)))
  cuts <- c(-1, cuts)
  
  varC <- cut(varCol, cuts)
  appC <- cut(appCol, cuts)
  
  # Now apply the categorical make prediction function
  make_prediction_categorical(outCol, varC, appC)
  
}

# set `did_churn` column to 0 or 1
subs <- subs %>% 
  mutate(did_churn = ifelse(TRUE, 1, 0))
```

We'll loop through the categorical variables and make predictions. Once the predictions are made, we calculate AUC on the training and testing sets.

```{r}
# identify categorical varaibles
catVars <- c('signup_option', 'is_mobile_user', 'billing_interval', 'country', 'simplified_plan_name',
             'signup_client_name')

# loop through categorical varaibles and mkae predictions
for(v in catVars) {
  
  # Make prediction for each categorical variable
  pi <- paste('pred', v, sep='_')
  
  # Do it for the training and testing datasets
  training[, pi] <- make_prediction_categorical(training[, "did_churn"], 
                                               training[, v], training[,v]) 
  
  testing[, pi] <- make_prediction_categorical(training[, "did_churn"], 
                                               training[, v], testing[,v]) 
}
```

For each level of the categorical levels, we have the predicted likelihood of a user churning. This is just the proportion of users that churned in each segment.

```{r}
# calculate AUC for each
for(v in catVars) {
  
  # Name the prediction variables
  pi <- paste('pred', v, sep = '_')
  
  # Find the AUC of the variable on the training set
  aucTrain <- calcAUC(training[, pi], training[, "did_churn"])
  
  # Find the AUC of the variable on the testing set  
  aucTest <- calcAUC(testing[, pi], testing[, "did_churn"])
    
  # Print the results
  print(sprintf("%s, trainAUC: %4.3f testingAUC: %4.3f", pi, aucTrain, aucTest))
    
}
```

These single variable models aren't any better than random guesses unfortunately. Let's loop through the numeric variables now. The predictions are made by essentially turning the numeric variables into categorical variables by creating "bins". There are a lot of numeric variables.

```{r message = F, warning = F}
# define numeric variables
numVars <- c('number_of_profiles', 'number_of_twitter_profiles', 'number_of_facebook_personal_profiles',
             'number_of_facebook_pages', 'number_of_facebook_groups', 'number_of_instagram_personal_profiles',
             'number_of_instagram_business_profiles', 'number_of_successful_charges', 
             'number_of_failed_charges', 'number_of_refunded_charges', 'days_since_last_update',
             'weeks_with_updates', 'total_updates', 'updates_per_week', 'estimate', 'subscription_age')

# loop through the columns and apply the formula
for(v in numVars) {
  
  # name the prediction column
  pi <- paste('pred', v, sep = '_')
  
  # Make predictions
  training[, pi] <- make_prediction_numeric(training[, "did_churn"], 
                                               training[, v], training[,v]) 
  
  testing[, pi] <- make_prediction_numeric(training[, "did_churn"], 
                                               training[, v], testing[,v]) 
  
  
 # Find the AUC of the variable on the training set
  aucTrain <- calcAUC(training[, pi], training[, "did_churn"])
  
  # Find the AUC of the variable on the testing set  
  aucTest <- calcAUC(testing[, pi], testing[, "did_churn"])
    
  # Print the results
  print(sprintf("%s, trainAUC: %4.3f testingAUC: %4.3f", pi, aucTrain, aucTest))
    
}
```

All of these single variable models perform very poorly. None do any better than random guessing. It seems like the variance in the response cannot be explained sufficiently by any single variable in our dataset. Makes sense.